---
title: "Reproducible Research: Peer Assessment 1"
author: "George Wangolo"
date: "February 9, 2016"
output: html_document
---

Introduction

The goal of this assignment to practice skills needed for reproducible research. Specifically this assignment use R markdown to write a report that answers the questions detailed in the sections below. In the process, the single R markdown document will be processed by knitr and be transformed into an HTML file.

Start this assignment with fork/clone the GitHub repository created for this assignment. When finish, submit this assignment by pushing your completed files into your forked repository on GitHub. The assignment submission will consist of the URL to your GitHub repository and the SHA-1 commit ID for your repository state.

Data

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute intervals through out the day. The data consists of two months of data from an anonymous individual collected during the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day. The data for this assignment can be downloaded from the course web site: Dataset: Activity monitoring data [52K] The variables included in this dataset are:

steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)
date: The date on which the measurement was taken in YYYY-MM-DD format
interval: Identifier for the 5-minute interval in which measurement was taken The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.
Loading and preprocessing

## Loading and preprocessing the data from a csv file from the current R working directory.

```{r, echo=TRUE}
activity <- read.csv("~/Desktop/Coursera/Reproducible Research/activity.csv", colClass=c('integer', 'Date', 'integer'))
```

## What is mean total number of steps taken per day?

### The total number of steps taken per day

```{r, echo=TRUE}
steps.date <- aggregate(steps ~ date, activity, sum)
head(steps.date)
```


Plot histogram of the total number of steps taken each day:

```{r, echo=TRUE}
barplot(steps.date$steps, names.arg=steps.date$date, ylim=c(0, 25000), 
        xlab="date", ylab="sum(steps)")
```

Report mean and median of steps:

Mean of total number of steps taken per day

```{r, echo=TRUE}
mean(steps.date$steps)
```

Median of total number of steps taken per day
```{r, echo=TRUE}
median(steps.date$steps)
```


## What is the average daily activity pattern?

Time series plot of the 5-minute interval and average number of steps taken averaged across all days

```{r, echo=TRUE}
steps.interval <- aggregate(steps ~ interval, activity, mean)
plot(steps.interval, type='l')
```

The 5-minute interval contains the maximum number of steps

```{r, echo=TRUE}
steps.interval$interval[which.max(steps.interval$steps)]
```


## Imputing missing values

The total number of missing values in the dataset is

```{r, echo=TRUE}
sum(is.na(activity$steps))
```

The strategy for filling in all of the missing values in the dataset is to use mean of the day.

Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r, echo=TRUE}
activity.clean <- merge(activity, steps.date, by="date", suffixes=c("", ".mean"))
nas <- is.na(activity.clean$steps)
activity.clean$steps[nas] <- activity.clean$steps.mean[nas]
activity.clean <- activity.clean[, c(1:3)]
head(activity.clean)
```

Histogram of the total number of steps taken each day

```{r, echo=TRUE}
steps.date <- aggregate(steps ~ date, activity.clean, sum)
barplot(steps.date$steps, names.arg=steps.date$date, ylim=c(0, 25000), 
        xlab="date", ylab="sum(steps),")
```

Mean of total number of steps taken per day

```{r, echo=TRUE}
mean(steps.date$steps)
```

Median of total number of steps taken per day

```{r, echo=TRUE}
median(steps.date$steps)
```
Data don’t appear to be significant different because imputation uses mean for that particular day but steps are NA for that entire day.

## Are there differences in activity patterns between weekdays and weekends?

Add new factor variable dayType with 2 levels – “weekday” and “weekend”

```{r, echo=TRUE}
dayType <- function(dates) {
  f <- function(date) {
    if (weekdays(date) %in% c("Saturday", "Sunday")) {
      "weekend"
    }
    else {
      "weekday"
    }
  }
  sapply(dates, f)
}

activity$dayType <- as.factor(dayType(activity$date))
str(activity)
```

Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekdays and weekends

```{r, echo=TRUE}
library(lattice)

steps.interval <- aggregate(steps ~ interval + dayType, activity, mean)
xyplot(steps ~ interval | dayType, data=steps.interval, layout=c(2,1), type='l')
```

